---
title: "TitanicDataset"
author: "Pradyumna Janga"
date: "August 25, 2017"
output: html_document
---
================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, message=FALSE, warning=FALSE, packages}
knitr::opts_chunk$set(echo = TRUE)
# Load all of the packages that you end up using
# in your analysis in this code chunk.
library(ggplot2)
library(plyr)
# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.
```

## Titanic Dataset

You are a Data Analyst working for White Star Line (the owners of RMS Titanic). The company is trying to better understand the passengers who were on board the ship when it sank. You have been supplied with a data set of passengers and head office would like to understand the following

1. How many passengers are included in the data?
2. Who bought the most expensive tickets?
3. How many men and how many women survived? (please provide a bar chart)
4. How many passengers are included in the data by class? (please provide a bar chart)
5. How many people had the title 'Mrs'?
6. What is the correlation between Age and Fare? (please provide a scatter plot)
7. What is the distribution of fares? (please provide a histogram)

We will first load the data into a variable

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
# Load the Data
setwd("C:/Users/vj853t/Documents/DataScience/Case Study/Case Study 6")
titanic_data <- read.csv("~/DataScience/Case Study/Case Study 6/titanic_data.csv")
colnames(titanic_data) = c('passanger_id', 'survived_status', 'passanger_class', 'name', 'sex', 'age', 'number-sibiling_spouse', 'parch_staus', 'ticket', 'fare', 'cabin', 'embarked')
```
The dataset consists of all the details of the passangers who travelled in Titanic ship. There are a total of 891 observations with 12 variables.

##Dataset Cleanup

Let us check the structure of the dataset

```{r echo=FALSE, message=FALSE, warning=FALSE, Summary}
str(titanic_data)
```

Passanger Class is an insteger carying from 1 to 3. To make it much better to read, we will rename 1 as 1st Class, 2 as 2nd Class and 3 as 3rd Class

```{}
titanic_data$passanger_class[titanic_data$passanger_class == 1] = '1st Class'
titanic_data$passanger_class[titanic_data$passanger_class == 2] = '2nd Class'
titanic_data$passanger_class[titanic_data$passanger_class == 3] = '3rd Class'
```

Let us check the class of Passanger Class
```{r echo=FALSE, message=FALSE, warning=FALSE, ClassCheck}
class(titanic_data$passanger_class)
```

Since there are 3 levels of classes, it is better to change it's class to factor instead of integer.
```{r echo=FALSE, message=FALSE, warning=FALSE, ClassChange}
titanic_data$passanger_class <- as.factor(titanic_data$passanger_class)
class(titanic_data$passanger_class)
```

=====================================================

##Data Exploration

Now that the passanger class is changed to factor, let us explore data further

There are a total of 891 members who travelled in Titanic. Let us see who bought the top 10 costliest tickets

```{r echo=FALSE, message=FALSE, warning=FALSE, ExpensiveTickets}
expensive_tickets <- head(titanic_data[order(titanic_data$fare, decreasing= T),], n=10)
expensive_tickets
```

3 people bought the costliest tickets each for $512.33, 4 people bought the 2nd costliest tickets each for $263.00 and 3 people 

Let us try to classify the number of people who travelled based on the passanger class they travelled.
```{r echo=FALSE, message=FALSE, warning=FALSE, classsummary}
survivors_class = aggregate(titanic_data$passanger_class, by = list(class = titanic_data$passanger_class), FUN = length)
survivors_class

ggplot(survivors_class, aes(x=class, y=x)) + geom_bar(stat="identity") + 
  labs(x="Passenger Travel Class", y="Number of Survivors") + ggtitle("Survivors by Class")
```

From the graph we draw insight about the fact that 491 members travelled in 3rd Class, 216 members travelled in 1st Class and 184 members travelled in 2nd Class. The least amount of people travelled in 2nd class and the most people travelled in 3rd class.

Among the people who survived, let us compare the gender.
```{r echo=FALSE, message=FALSE, warning=FALSE, gendersummary}
survivors_sex = aggregate(titanic_data$survived_status, by = list(sex = titanic_data$sex), FUN = sum)
colnames(survivors_sex) <- c('Sex', 'Count')
ggplot(survivors_sex, aes(Sex, Count)) + geom_bar(stat="identity") + 
  labs(x="Sex", y="Number of Survivors") + ggtitle("Number of Survivors by Sex")
```

Out of 342 people who survived the wreck, there are 233 women and 109 men. Twice the amount of women than men have survived. So, the ratio of men is to women who survived titanic is 1:2.

Out of the 233 women who survived the titanic, let us see how many of them are married. 
```{r echo=FALSE, message=FALSE, warning=FALSE, marriedwomen}
nrow(titanic_data[grepl('Mrs', titanic_data$name),])
```

Out of 233 women who survived, 129 of them were married women who make up 55%.

Now, check if there is any correlation between the age and fare.

```{r echo=FALSE, message=FALSE, warning=FALSE, agefarecorrelation}
cor(titanic_data$age, titanic_data$fare, use='complete.obs')
plot(titanic_data$age, titanic_data$fare, xlab = "Age of Passengers", ylab ="Ticket Fare", main = "Correlation between Age and Fare")
```

The correlation factor of Fare and Age is 0.09 which is minimal. That indicates that the fare has nothing to fo with the age of the passengers. It also indicates that age does not guarantee financial success. Both are dependent only to a smaller extent which means not everyone knows how to earn more money as they grow old. 







